---
title: "BIOS 640  Assignment Week 3"
author: "Emmanuella"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
  )

library(tidyverse)
library(cowplot)    
library(ggpubr)      

if (!dir.exists("figures")) dir.create("figures")
```

```{r load-data}
nhanes <- read_csv("cleaned_NHANES.csv") %>%
  select(-`...1`) 

diet <- read_csv("diet.csv")
```

# Exercise 1: Reproducing and arranging ggplot2 figures {.tabset}
## Individual Figures

```{r ex1-fig1, fig.width=7, fig.height=4}
# Figure 1: distribution of age, split by gender, in 5-year bins
fig1 <- ggplot(nhanes, aes(x = age, fill = gender)) +
  geom_histogram(
    binwidth = 5,
    boundary = 0,                                # bins start at multiples of 5
    position = position_dodge2(preserve = "single")
  ) +
  labs(
    x = "Age (years)",
    y = "Count",
    fill = "Gender"
  )

fig1
```

```{r ex1-fig2, fig.width=7, fig.height=4}
# Figure 2: proportion of gender within each ethnicity category (RIDRETH3 version)
fig2 <- ggplot(nhanes, aes(x = ethnicity_2, fill = gender)) +
  geom_bar(position = "fill") +
  labs(
    x = "Ethnicity",
    y = "Proportion",
    fill = "Gender"
  )

fig2
```

## Combining the two figures

```{r ex1-cowplot, fig.width=7, fig.height=7}
# Method 1: cowplot::plot_grid()
plot_grid(
  fig1, fig2,
  ncol = 1,
  labels = c("A", "B"),
  align = "v"
)
```

```{r ex1-ggarrange, fig.width=7, fig.height=7}
# Method 2: ggpubr::ggarrange()
ggarrange(
  fig1, fig2,
  ncol = 1, nrow = 2,
  labels = c("A", "B"),
  common.legend = TRUE,
  legend = "right"
)
```

## Comparing `plot_grid()` and `ggarrange()`

Both functions accomplish the same basic goal of stitching separate `ggplot`
objects into a single output, but they make different trade-offs for this
particular pair of figures. `plot_grid()` from **cowplot** keeps each plot's
own legend, which here means the "Gender" legend is printed twice — once
per panel — even though it is identical both times. This is redundant and
uses vertical space that could otherwise go to the plots themselves.
`ggarrange()` from **ggpubr**, on the other hand, supports
`common.legend = TRUE`, which collapses the two identical legends into a
single shared legend placed to the side of the whole arrangement. For these
two figures specifically, that makes `ggarrange()` the better layout: the
panels are taller, the duplicated legend clutter is gone, and it is still
immediately obvious which color corresponds to which gender in both plots.
`plot_grid()` would be the more attractive option if the two figures had
different fill variables (so that each legend carried unique information),
but that isn't the case here.

# Exercise 2: Visualizing key characteristics

library(DT)
datatable(nhanes)

```{r ex2-age, fig.width=6, fig.height=4}
age_plot <- ggplot(nhanes, aes(x = age)) +
  geom_histogram(binwidth = 5, boundary = 0, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Age",
    x = "Age (years)",
    y = "Count"
  )

age_plot
ggsave("figures/age_distribution.png", age_plot, width = 6, height = 4)
```

**Description:** Age is right-skewed toward younger participants. The
largest 5-year bin is 0-4 years (`r sum(nhanes$age < 5)` participants), and
counts decline steadily through childhood and adolescence before leveling
off to a roughly flat plateau of about 1,300-1,700 participants per 5-year
bin across adulthood (ages 20-70), then tapering off after age 70 — NHANES
top-codes age at 80. There are
`r sum(is.na(nhanes$age))` missing values for age.

```{r ex2-gender, fig.width=5, fig.height=4}
gender_plot <- ggplot(nhanes, aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Distribution of Gender",
    x = "Gender",
    y = "Count",
    fill = "Gender"
  )

gender_plot
ggsave("figures/gender_distribution.png", gender_plot, width = 5, height = 4)
```

**Description:** The sample is close to evenly split between genders, with
`r sum(nhanes$gender == "Female")` females and
`r sum(nhanes$gender == "Male")` males — a difference of less than 2%.
There are `r sum(is.na(nhanes$gender))` missing values for gender.

```{r ex2-ethnicity1, fig.width=6.5, fig.height=4}
ethnicity1_plot <- ggplot(nhanes, aes(x = fct_infreq(ethnicity_1), fill = ethnicity_1)) +
  geom_bar() +
  labs(
    title = "Distribution of Ethnicity (5-category version)",
    x = "Ethnicity",
    y = "Count",
    fill = "Ethnicity"
  )

ethnicity1_plot
ggsave("figures/ethnicity1_distribution.png", ethnicity1_plot, width = 6.5, height = 4)
```

**Description:** The 5-category ethnicity variable is dominated by the
White (n = `r sum(nhanes$ethnicity_1 == "White")`) and Black
(n = `r sum(nhanes$ethnicity_1 == "Black")`) groups, followed by Mexican
(n = `r sum(nhanes$ethnicity_1 == "Mexican")`), Other
(n = `r sum(nhanes$ethnicity_1 == "Other")`), and Other Hispanic
(n = `r sum(nhanes$ethnicity_1 == "Other Hispanic")`). There are
`r sum(is.na(nhanes$ethnicity_1))` missing values.

```{r ex2-ethnicity2, fig.width=6.5, fig.height=4}
ethnicity2_plot <- ggplot(nhanes, aes(x = fct_infreq(ethnicity_2), fill = ethnicity_2)) +
  geom_bar() +
  labs(
    title = "Distribution of Ethnicity (6-category version)",
    x = "Ethnicity",
    y = "Count",
    fill = "Ethnicity"
  )

ethnicity2_plot
ggsave("figures/ethnicity2_distribution.png", ethnicity2_plot, width = 6.5, height = 4)
```

**Description:** The 6-category ethnicity variable has the same White, Black,
and Mexican counts as the 5-category version, but it splits out an Asian
group (n = `r sum(nhanes$ethnicity_2 == "Asian")`) that was previously
folded into "Other," leaving a much smaller residual Other category
(n = `r sum(nhanes$ethnicity_2 == "Other")`). There are
`r sum(is.na(nhanes$ethnicity_2))` missing values.

## Comparing the two ethnicity variables

`ethnicity_1` (originally `RIDRETH1`) and `ethnicity_2` (originally
`RIDRETH3`) agree on the White, Black, Mexican American, and Other Hispanic
categories, but `ethnicity_2` separately identifies Asian participants
(n = `r sum(nhanes$ethnicity_2 == "Asian")`) instead of lumping them into a
catch-all "Other" group. `ethnicity_1`'s "Other" category is therefore
larger and less informative (n = `r sum(nhanes$ethnicity_1 == "Other")` vs.
n = `r sum(nhanes$ethnicity_2 == "Other")`), since it mixes Asian
participants together with everyone who did not fit the other four labels.
I will keep **`ethnicity_2`** because it provides more granular, more
accurate self-reported race information without losing any of the detail
present in `ethnicity_1`, and it reflects the newer NHANES coding scheme
(`RIDRETH3`) that replaced `RIDRETH1` in more recent survey cycles.

```{r ex2-drop-ethnicity1}
nhanes <- nhanes %>% select(-ethnicity_1)
```

# Exercise 3: Improving a ggplot figure

## Critique of the provided figure

My colleague's figure plots raw body weight for all 20 participants over the
16-week study, with each participant assigned a distinct color from
ggplot2's default palette. Several things make it hard to actually answer
the study question -- "how did each participant's weight change relative to
their own baseline?":

1. **Twenty colors is too many to read.** The default hue palette recycles
   similar shades (several near-identical blues and teals, e.g. participants
   10-14), so adjacent lines are nearly indistinguishable and the legend
   itself becomes a long, hard-to-scan list rather than a useful key.
2. **The y-axis shows absolute weight, not change from baseline.** Because
   participants start at very different weights (roughly 65 kg to 100 kg),
   a participant who started heavy will always sit higher on the chart even
   if they lost more weight than a lighter participant. The figure cannot
   directly answer the study's actual question of *change* over time.
3. **A genuinely interesting pattern is buried by the clutter.** Several
   participants' weight lines are perfectly flat across all 16 weeks (no
   change at all), while others clearly decrease. Because of the
   overplotting and the 20-color legend, this distinction is easy to miss
   entirely, even though it is one of the most important findings in the
   data.
4. **The default grey theme and gridlines add visual noise** without adding
   information, competing with the 20 data lines for attention.

## Proposed improvements

To fix these issues, I will re-express weight as **change from each
participant's own baseline weight (Week 0)**, so every line starts at zero
and the y-axis directly shows weight lost or gained. I will also replace the
20-line color legend with a **3-level outcome group** (weight loss / no
change / weight gain, based on each participant's total change over the
study), which cuts the palette down to three easily distinguishable colors
and immediately communicates the overall pattern in the data. Finally, I
will add a horizontal reference line at zero (the baseline) and switch to a
cleaner, non-default theme with clear titles and axis labels.

```{r ex3-prep}
diet <- diet %>%
  group_by(Participant) %>%
  mutate(
    baseline = Weight[Week == 0],
    weight_change = Weight - baseline
  ) %>%
  ungroup()

# Classify each participant by their total change over the study
outcome_summary <- diet %>%
  group_by(Participant) %>%
  summarize(total_change = weight_change[Week == max(Week)], .groups = "drop") %>%
  mutate(
    outcome_group = case_when(
      total_change < -0.5 ~ "Weight loss",
      total_change >  0.5 ~ "Weight gain",
      TRUE               ~ "No change"
    ),
    outcome_group = factor(
      outcome_group,
      levels = c("Weight loss", "No change", "Weight gain")
    )
  )

diet <- diet %>%
  left_join(outcome_summary %>% select(Participant, outcome_group), by = "Participant")
```

```{r ex3-improved-figure, fig.width=7, fig.height=4.5}
diet_fig <- ggplot(diet, aes(x = Week, y = weight_change, group = Participant, color = outcome_group)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(alpha = 0.8, linewidth = 0.7) +
  scale_color_manual(
    values = c(
      "Weight loss" = "#1b9e77",
      "No change"   = "#7570b3",
      "Weight gain" = "#d95f02"
    )
  ) +
  labs(
    title = "Change in Body Weight Relative to Baseline (Week 0)",
    x = "Week",
    y = "Weight Change from Baseline (kg)",
    color = "Outcome group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

diet_fig
```

This version makes the study's actual finding immediately visible: most
participants lost a substantial amount of weight over the 16 weeks, a
handful showed no change at all, and a few actually gained weight. Every
line starts at zero by construction, so the vertical position of a line at
any week directly represents that participant's progress -- something the
original figure could not show.


# Exercise 4: Exploring relationships between age, gender, and blood pressure

Blood pressure was recorded up to four times per participant in this survey.
To get a single, stable measure of each person's systolic blood pressure
(SBP), I averaged their available readings, requiring a value to be missing
only when *all four* readings were missing rather than dropping anyone with
a single missing reading.

```{r ex4-avg-sbp}
nhanes <- nhanes %>%
  rowwise() %>%
  mutate(
    avg_sbp = if (all(is.na(c_across(starts_with("systolic_bp"))))) {
      NA_real_
    } else {
      mean(c_across(starts_with("systolic_bp")), na.rm = TRUE)
    }
  ) %>%
  ungroup()
```

Restricting the sample to participants with complete information on age,
gender, and this new average SBP variable brings the analytic sample from
`r nrow(nhanes)` participants down to a final sample described below.

```{r ex4-filter}
nhanes_final <- nhanes %>%
  filter(!is.na(age), !is.na(gender), !is.na(avg_sbp))
```

Almost all of that drop comes from missing blood pressure data: age and
gender have no missing values in this data set, so every participant
excluded from the final sample (`r nrow(nhanes) - nrow(nhanes_final)` of the
original `r nrow(nhanes)`) was excluded because none of their four SBP
readings were recorded.

With average SBP in hand, I grouped participants into the standard clinical
categories used to describe blood pressure, keeping the categories in their
natural clinical order (Normal < Elevated < Stage 1 hypertension < Stage 2
hypertension) so that later plots and legends read in a meaningful sequence
rather than alphabetically.

```{r ex4-hypertension-cat}
nhanes_final <- nhanes_final %>%
  mutate(
    hypertension_cat = case_when(
      avg_sbp < 120 ~ "Normal",
      avg_sbp >= 120 & avg_sbp < 130 ~ "Elevated",
      avg_sbp >= 130 & avg_sbp < 140 ~ "Stage 1 hypertension",
      avg_sbp >= 140 ~ "Stage 2 hypertension"
    ),
    hypertension_cat = factor(
      hypertension_cat,
      levels = c("Normal", "Elevated", "Stage 1 hypertension", "Stage 2 hypertension"),
      ordered = TRUE
    ),
    age_cat = factor(
      age_cat,
      levels = c("0-17", "18-24", "25-39", "40-59", "60+"),
      ordered = TRUE
    )
  )
```

## Describing the final sample

Four plots below summarize who remains in the analytic sample and how it
compares to the full data set. First, the sample-size plot makes the size of
the drop concrete: the final sample retains
`r round(100 * nrow(nhanes_final) / nrow(nhanes), 1)`% of the original
participants.

```{r ex4-plot-samplesize, fig.width=4.5, fig.height=4}
sample_size_df <- tibble(
  sample = factor(c("Full sample", "Final sample"), levels = c("Full sample", "Final sample")),
  n = c(nrow(nhanes), nrow(nhanes_final))
)

samplesize_plot <- ggplot(sample_size_df, aes(x = sample, y = n, fill = sample)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4) +
  labs(
    title = "Sample Size Before and After Filtering",
    x = NULL,
    y = "Number of participants"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

samplesize_plot
```

Next, the distribution of hypertension categories in the final sample shows
that a majority of participants have Normal average SBP, with the remaining
participants fairly evenly spread across the Elevated and two hypertension
stages.

```{r ex4-plot-hypertension-counts, fig.width=5.5, fig.height=4}
hyp_counts_plot <- ggplot(nhanes_final, aes(x = hypertension_cat, fill = hypertension_cat)) +
  geom_bar() +
  labs(
    title = "Hypertension Category Counts (Final Sample)",
    x = "Hypertension category",
    y = "Count",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

hyp_counts_plot
```

Finally, I compare gender in the full sample against gender in the final
sample by reusing the exact plot object I saved in Exercise 2
(`gender_plot`) alongside a newly built version restricted to the final
sample.

```{r ex4-plot-gender-compare, fig.width=9, fig.height=4}
gender_final_plot <- ggplot(nhanes_final, aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Distribution of Gender (Final Sample)",
    x = "Gender",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal()

plot_grid(
  gender_plot + labs(title = "Full Sample") + theme_minimal(),
  gender_final_plot + labs(title = "Final Sample"),
  ncol = 2,
  align = "h"
)
```

The gender balance barely shifts after filtering: both the full and final
samples remain close to an even split between Male and Female participants,
which tells us that missingness in blood pressure was not meaningfully
related to gender. Combined with the sample-size plot above, this supports
treating the final sample as reasonably representative of the full data set
on this dimension, even though it lost close to a quarter of its rows.

## How do hypertension categories vary by age and gender?

To explore the relationship between age, gender, and blood pressure
together, I faceted hypertension category counts by gender across the
(already ordered) age categories. Categories are ordered from youngest to
oldest along the x-axis and from Normal to Stage 2 hypertension in the
legend, a custom color palette moves from a reassuring green through to a
warning red as severity increases, and a non-default theme keeps the panels
clean and easy to compare side by side.

```{r ex4-facet-plot, fig.width=8, fig.height=5}
facet_plot <- ggplot(nhanes_final, aes(x = age_cat, fill = hypertension_cat)) +
  geom_bar(position = "fill") +
  facet_wrap(~ gender) +
  scale_fill_manual(
    values = c(
      "Normal" = "#2ca25f",
      "Elevated" = "#fee08b",
      "Stage 1 hypertension" = "#fc8d59",
      "Stage 2 hypertension" = "#d73027"
    )
  ) +
  labs(
    title = "Blood Pressure Category by Age Group and Gender",
    x = "Age group",
    y = "Proportion of participants",
    fill = "Hypertension category"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey85")
  )

facet_plot
```

The pattern across age groups is the clearest story in this data set: the
proportion of participants with Normal average SBP shrinks steadily from
childhood through older adulthood, while Stage 1 and Stage 2 hypertension
become progressively more common, with the oldest group (60+) showing by
far the largest share of Stage 2 hypertension in both panels. The two facets
-- Male and Female -- follow this same age gradient, but the male panel
shows a visibly larger hypertension share in the middle-aged groups
(40-59) than the female panel does at the same ages, a gap that narrows
again in the 60+ group. Taken together, age is the dominant driver of
blood pressure category in this sample, with gender acting as a secondary
factor that shifts when, in the life course, elevated blood pressure
becomes common rather than changing the overall upward trend with age.
